---
title: "Air Pollution and Health Outcomes in the UK Biobank"
output:
  github_document:
    toc: true
    toc_depth: 4
author: "Caitlyn Chitwood"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
  condition:
    label: "Select Condition (LC = Lung Cancer, CH = Clonal Hematopoiesis)"
    value: "LC"
    input: select
    choices: ["LC", "CH"]
  printcode:
    label: "Display Code:"
    value: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = params$printcode, out.width = 4, out.height = 4)

libraries <- c("survival", "ggplot2", "rstudioapi", "readr", "dplyr", "data.table",
               "shiny", "knitr")
lapply(libraries, require, character.only = TRUE) #load multiple packages

source("utils.R")  #load functions
source("plots.R")  #load functions
source("data_process.R")  #load functions

```

```{r data-processing, include = FALSE}

df_pheno <- get_data(params$condition) #generate data frame for selected data
attach(df_pheno)

```

`r if(params$condition=="LC"){"Lung cancer"} else if(params$condition=="CH"){"Clonal hematopoiesis"}` data was used for the following analyses.

## Descriptive Statistics

### Summary Statistics

```{r summary}
lungcancer <- table(lung_cancer)

kable(head(df_pheno),
      col.names = TRUE,
      caption = "A table showing first 6 rows of relevant variables.",
      )
```

There are `r lungcancer[2]` cases of lung cancer in the data set.

### Density Plots

```{r density-plot-sex}
kd_male <- density(na.omit(pm10_2007[sex=="Male"])) #define kernel density
plot(kd_male,
     col='blue',
     lwd=2,
     main = "Female/Male PM10_2007") #create kernel density plot
kd_female <- density(na.omit(pm10_2007[sex=="Female"]))
lines(kd_female,
      col='red',
      lwd=2)

# fill in kernel density plot with specific color
# polygon(kd_male, col='blue', border='black')
# polygon(kd_female, col='red', border='black')
```

```{r density-plots}
x1 = na.omit(pm10_2007[ageBaseline>=60])
x2 = na.omit(pm10_2007[ageBaseline<60])
plot_mult_density("pm10_2007 (blue=60 or over, red=under 60)", x1,x2)
```

### Histograms

```{r histogram-age}
#define data
x1 = na.omit(pm10_2007[ageBaseline>=60])
x2 = na.omit(pm10_2007[ageBaseline<60])

#plot two histograms in same graph
hist(x1, col='red', main='pm10_2007', xlab='(μg/m3)')
hist(x2, col='green', add=TRUE)

#add legend
legend('topright', c('60 or Over', 'Under 60'), fill=c('red', 'green'))
```

```{r histogram}
hist(pm25_2010, 
     main="Particulate matter air pollution (pm2.5) 2010",
     xlab="PM2.5 (μg/m3)",
     col="cadetblue",
     freq=FALSE,
     ylim=c(0,0.7))

#Plot lognormal distribution
xfit <- seq(min(pm25_2010, na.rm=TRUE), max(pm25_2010, na.rm=TRUE), length=40)
yfit <- dlnorm(xfit,meanlog=log(mean(pm25_2010)), sdlog=log(sd(pm25_2010)))
lines(xfit, yfit, col="black", lwd=2)

```

The distribution is right-skewed and appears to be a log normal distribution.

```{r histogram2, fig.height=7, fig.width=7}
par(mfrow=c(3,3))
hist(pm10_2007, 
     xlab="pm10_2007 (μg/m3)",
     col="cadetblue",
     freq=FALSE)

hist(pm10_2010, 
     xlab="pm10_2010 (μg/m3)",
     col="cadetblue",
     freq=FALSE)

hist(pmcourse_2010, 
     xlab="pmcourse_2010 (μg/m3)",
     col="cadetblue",
     freq=FALSE)

hist(no2_2005, 
     xlab="no2_2005 (μg/m3)",
     col="cadetblue",
     freq=FALSE)

hist(no2_2006, 
     xlab="no2_2006 (μg/m3)",
     col="cadetblue",
     freq=FALSE)

hist(no2_2007, 
     xlab="no2_2007 (μg/m3)",
     col="cadetblue",
     freq=FALSE)

hist(no2_2010, 
     xlab="no2_2010 (μg/m3)",
     col="cadetblue",
     freq=FALSE)

hist(no1_2010, 
     xlab="no1_2010 (μg/m3)",
     col="cadetblue",
     freq=FALSE)

hist(pm25_2010, 
     xlab="PM2.5 (μg/m3)",
     col="cadetblue",
     freq=FALSE)
```

### Scatter Plots and Correlation Coefficients: pollutants

```{r scatterplot-PM10-NO2}
plot(pm10_2010, no2_2010)
cor(pm10_2010, no2_2010, method="pearson", use="complete.obs")
cor(pm10_2010, no2_2010, method="spearman", use="complete.obs")
```

```{r scatterplot-PM10-PM2.5}
plot(pm10_2010,pm25_2010)
cor(pm10_2010, pm25_2010, method="pearson", use="complete.obs")
cor(pm10_2010, pm25_2010, method="spearman", use="complete.obs")
```

```{r scatterplot-NO1-NO2}
plot(no1_2010, no2_2010)
cor(no1_2010, no2_2010, method="pearson", use="complete.obs")
cor(no1_2010, no2_2010, method="spearman", use="complete.obs")
```

### Scatter Plots and Correlation Coefficients: pollutants and covariates

```{r correlation}
cor(pm25_2010, ageBaseline, use="pairwise.complete.obs")
cor(pm25_2010, bmi, use="pairwise.complete.obs")
cor(pm25_2010, packYear, use="pairwise.complete.obs")
```

### PM 2.5

#### Box Plot

```{r boxplot}
quantile(pm25_2010, na.rm=TRUE)
iqr <- IQR(pm25_2010, na.rm=TRUE)

boxplot(pm25_2010, 
        horizontal=TRUE,
        col="cadetblue")
```

#### QQ Plot

```{r qqplot}
qqnorm(pm25_2010)
qqline(pm25_2010)
```

#### Outliers

```{r outliers}
extremeOut <- 3*iqr
mildOut <- 1.5*iqr

extremePM25 <- subset(df_pheno, pm25_2010 > (mean(pm25_2010, na.rm=TRUE)+ extremeOut))
mildPM25 <- subset(df_pheno, pm25_2010 > (mean(pm25_2010, na.rm=TRUE)+ mildOut))
```

There are `r nrow(extremePM25)` extreme outliers and `r nrow(mildPM25)` mild outliers for PM2.5 data from 2010.

#### Correlation Coefficients: pollutant extreme outliers and covariates

```{r correlation-extreme}
cor(extremePM25$pm25_2010, extremePM25$ageBaseline)
cor(extremePM25$pm25_2010, extremePM25$bmi, use="pairwise.complete.obs")
cor(extremePM25$pm25_2010, extremePM25$packYear, use="pairwise.complete.obs")
```

#### Correlation Coefficients: pollutant mild outliers and covariates

```{r correlation-mild}
cor(mildPM25$pm25_2010, mildPM25$ageBaseline, use="pairwise.complete.obs")
cor(mildPM25$pm25_2010, mildPM25$bmi, use="pairwise.complete.obs")
cor(mildPM25$pm25_2010, mildPM25$packYear, use="pairwise.complete.obs")

rm(mildPM25, extremePM25)
```

## Regression: Cox Proportional-Hazards Model

### Model 1: age and sex covariates only

```{r model-1}
res.model1 <- coxph(
  Surv(t_lungCancer, lung_cancer) ~  pm25_2010per5 + ageBaseline + sex,
  data = df_pheno,
  ties = "efron"
)
res.model1
```

### Model 2: multiple covariates

```{r model-2}
res.model2 <- coxph(
  Surv(t_lungCancer, lung_cancer) ~ pm25_2010per5 + ageBaseline + sex + householdIncomeCat + education + bmi + smokingCat +  exposeSmokeHomeRaw,
  data = df_pheno,
  ties = "efron"
)
res.model2
```

## Citations

[@huang2021a; @tian2022a; @swanton2016]
