---
title: "Air Pollution and Health Outcomes in the UK Biobank"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
author: "Caitlyn Chitwood - "
date: "`r format(Sys.time(), '%B %d, %Y')`"
params:
  condition:
    label: "Select Condition (LC = Lung Cancer, CH = Clonal Hematopoiesis)"
    value: "CH"
    input: select
    choices: ["LC", "CH"]
  printcode:
    label: "Display Code:"
    value: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = params$printcode)

libraries <-
  c(
    "survival",
    "ggplot2",
    "readr",
    "dplyr",
    "data.table",
    "shiny",
    "knitr",
    "summarytools",
    "rstudioapi",
    "sjPlot",
    "writexl",
    "anytime",
    "table1"
  )
lapply(libraries, require, character.only = TRUE) #load multiple packages

source("utils.R")  #load functions
source("plots.R")  #load functions
source("data_process.R")  #load functions

eval_LC <- params$condition == "LC"
eval_CH <- params$condition == "CH"
```

```{r data-processing, include = FALSE}

df_pheno <- get_data(params$condition) #generate data frame for selected data
attach(df_pheno)

```

`r if(eval_LC){"Lung cancer"} else if(eval_CH){"Clonal hematopoiesis"}` data was used for the following analyses.


## Summary Statistics

There are `r if(eval_LC){sum(lung_cancer)} else if(eval_CH){sum(CH)}` cases in the data set.

```{r summary-table1}
table1(~ ageCat + sex + householdIncomeCat + eduCat + bmiCat + smokingCat, data=df_pheno, overall=c(left="Total"))
```

Average packages of cigarettes per day multiplied by years smoking: 
  `r mean(packYear, na.rm = TRUE)`

```{r summary-overall, results="asis"}
dfSummary(df_pheno, 
          round.digits = 2,
          plain.ascii  = FALSE, 
          style        = "grid", 
          graph.magnif = 0.65, 
          valid.col    = FALSE,
          tmp.img.dir  = "./tmp")

kable(head(df_pheno))
```

## Density Plots

```{r density-plot-sex}
kd_male <- density(na.omit(pm10_2007[sex=="Male"])) #define kernel density
plot(kd_male,
     col='blue',
     lwd=2,
     main = "Female/Male PM10_2007") #create kernel density plot
kd_female <- density(na.omit(pm10_2007[sex=="Female"]))
lines(kd_female,
      col='red',
      lwd=2)

# fill in kernel density plot with specific color
# polygon(kd_male, col='blue', border='black')
# polygon(kd_female, col='red', border='black')
```

```{r density-plots}
x1 = na.omit(pm10_2007[ageBaseline>=60])
x2 = na.omit(pm10_2007[ageBaseline<60])
plot_mult_density("pm10_2007 (blue=60 or over, red=under 60)", x1,x2)
```

## Histograms
### Pollutants

```{r histogram2, fig.height=7, fig.width=7}
par(mfrow=c(3,3))
hist(pm10_2007, 
     xlab="pm10_2007 (μg/m3)",
     col="cadetblue",
     freq=FALSE)

hist(pm10_2010, 
     xlab="pm10_2010 (μg/m3)",
     col="cadetblue",
     freq=FALSE)

hist(pmcoarse_2010, 
     xlab="pmcoarse_2010 (μg/m3)",
     col="cadetblue",
     freq=FALSE)

hist(no2_2005, 
     xlab="no2_2005 (μg/m3)",
     col="cadetblue",
     freq=FALSE)

hist(no2_2006, 
     xlab="no2_2006 (μg/m3)",
     col="cadetblue",
     freq=FALSE)

hist(no2_2007, 
     xlab="no2_2007 (μg/m3)",
     col="cadetblue",
     freq=FALSE)

hist(no2_2010, 
     xlab="no2_2010 (μg/m3)",
     col="cadetblue",
     freq=FALSE)

hist(no1_2010, 
     xlab="no1_2010 (μg/m3)",
     col="cadetblue",
     freq=FALSE)

hist(pm25_2010, 
     xlab="PM2.5 (μg/m3)",
     col="cadetblue",
     freq=FALSE)
```

## Scatter Plots
### Pollutants

```{r scatterplot-PM10-NO2}
plot(pm10_2010, no2_2010)
cor(pm10_2010, no2_2010, method="pearson", use="complete.obs")
cor(pm10_2010, no2_2010, method="spearman", use="complete.obs")
```

```{r scatterplot-PM10-PM2.5}
plot(pm10_2010,pm25_2010)
cor(pm10_2010, pm25_2010, method="pearson", use="complete.obs")
cor(pm10_2010, pm25_2010, method="spearman", use="complete.obs")
```

```{r scatterplot-NO1-NO2}
plot(no1_2010, no2_2010)
cor(no1_2010, no2_2010, method="pearson", use="complete.obs")
cor(no1_2010, no2_2010, method="spearman", use="complete.obs")
```

### Pollutants and covariates

```{r correlation}
cor(pm25_2010, ageBaseline, use="pairwise.complete.obs")
cor(pm25_2010, bmi, use="pairwise.complete.obs")
cor(pm25_2010, packYear, use="pairwise.complete.obs")
```

## Additional Plots
### PM 2.5

#### Box Plot

```{r boxplot}
quantile(pm25_2010, na.rm=TRUE)
iqr <- IQR(pm25_2010, na.rm=TRUE)

boxplot(pm25_2010, 
        horizontal=TRUE,
        col="cadetblue")
```

#### QQ Plot

```{r qqplot}
qqnorm(pm25_2010)
qqline(pm25_2010)
```

#### Outliers

```{r outliers}
extremeOut <- 3*iqr
mildOut <- 1.5*iqr

extremePM25 <- subset(df_pheno, pm25_2010 > (mean(pm25_2010, na.rm=TRUE)+ extremeOut))
mildPM25 <- subset(df_pheno, pm25_2010 > (mean(pm25_2010, na.rm=TRUE)+ mildOut))
```

There are `r nrow(extremePM25)` extreme outliers and `r nrow(mildPM25)` mild outliers for PM2.5 data from 2010.

#### Correlation Coefficients: pollutant extreme outliers and covariates

```{r correlation-extreme}
cor(extremePM25$pm25_2010, extremePM25$ageBaseline)
cor(extremePM25$pm25_2010, extremePM25$bmi, use="pairwise.complete.obs")
cor(extremePM25$pm25_2010, extremePM25$packYear, use="pairwise.complete.obs")
```

#### Correlation Coefficients: pollutant mild outliers and covariates

```{r correlation-mild}
cor(mildPM25$pm25_2010, mildPM25$ageBaseline, use="pairwise.complete.obs")
cor(mildPM25$pm25_2010, mildPM25$bmi, use="pairwise.complete.obs")
cor(mildPM25$pm25_2010, mildPM25$packYear, use="pairwise.complete.obs")

rm(mildPM25, extremePM25)
```

## Cox Regression

### Model 1: age and sex only

#### PM2.5 (2010)
```{r model-1-pm25-2010, eval=eval_LC}
res_model_1_pm25_2010 <- coxph(
  Surv(timeStudy, lung_cancer) ~ pm25_2010per5 + ageBaseline + sex,
  data = df_pheno,
  ties = "efron"
)
summary(res_model_1_pm25_2010)

sj_data = res_model_1_pm25_2010 %>% sjPlot::get_model_data(type="est")
sj_data
write_xlsx(sj_data, "master_files/model1/pm25-2010.xlsx")
```

### Model 2: multiple covariates

#### PM2.5 (2010)
```{r model-2-pm25-2010, eval=eval_LC}
res_model_2_pm25_2010 <- coxph(
  Surv(timeStudy, lung_cancer) ~ pm25_2010per5 + ageBaseline + sex + householdIncomeCat + eduCat + bmiCat + smokingCat + race,
  data = df_pheno,
  ties = "efron"
)
summary(res_model_2_pm25_2010)

sj_data = res_model_2_pm25_2010 %>% sjPlot::get_model_data(type="est")
sj_data
write_xlsx(sj_data, "master_files/model2/pm25-2010.xlsx")
```

#### PM10 (2010)
```{r model-2-pm10-2010, eval=eval_LC}
res_model_2_pm10_2010 <- coxph(
  Surv(timeStudy, lung_cancer) ~ pm10_2010per10 + ageBaseline + sex + householdIncomeCat + eduCat + bmiCat + smokingCat + race,
  data = df_pheno,
  ties = "efron"
)
summary(res_model_2_pm10_2010)

sj_data = res_model_2_pm10_2010 %>% sjPlot::get_model_data(type="est")
sj_data
write_xlsx(sj_data, "master_files/model2/pm10-2010.xlsx")
```

#### PM10 (2007)

```{r model-2-pm10-2007, eval=eval_LC}
res_model_2_pm10_2007 <- coxph(
  Surv(timeStudy, lung_cancer) ~ pm10_2007per10 + ageBaseline + sex + householdIncomeCat + eduCat + bmiCat + smokingCat + race,
  data = df_pheno,
  ties = "efron"
)
summary(res_model_2_pm10_2007)

sj_data = res_model_2_pm10_2007 %>% sjPlot::get_model_data(type="est")
sj_data
write_xlsx(sj_data, "master_files/model2/pm10-2007.xlsx")
```

#### PM Coarse (2010)

```{r model-2-pmcoarse-2010, eval=eval_LC}
res_model_2_pmcoarse_2010 <- coxph(
  Surv(timeStudy, lung_cancer) ~ pmcoarse_2010per5 + ageBaseline + sex + householdIncomeCat + eduCat + bmiCat + smokingCat + race,
  data = df_pheno,
  ties = "efron"
)
summary(res_model_2_pmcoarse_2010)

sj_data = res_model_2_pmcoarse_2010 %>% sjPlot::get_model_data(type="est")
sj_data
write_xlsx(sj_data, "master_files/model2/pmcoarse-2010.xlsx")
```


#### NO2 (2005)

```{r model-2-no2-2005, eval=eval_LC}
res_model_2_no2_2005 <- coxph(
  Surv(timeStudy, lung_cancer) ~ no2_2005per10 + ageBaseline + sex + householdIncomeCat + eduCat + bmiCat + smokingCat + race,
  data = df_pheno,
  ties = "efron"
)
summary(res_model_2_no2_2005)

sj_data = res_model_2_no2_2005 %>% sjPlot::get_model_data(type="est")
sj_data
write_xlsx(sj_data, "master_files/model2/no2-2005.xlsx")
```

#### NO2 (2006)

```{r model-2-no2-2006, eval=eval_LC}
res_model_2_no2_2006 <- coxph(
  Surv(timeStudy, lung_cancer) ~ no2_2006per10 + ageBaseline + sex + householdIncomeCat + eduCat + bmiCat + smokingCat + race,
  data = df_pheno,
  ties = "efron"
)
summary(res_model_2_no2_2006)

sj_data = res_model_2_no2_2006 %>% sjPlot::get_model_data(type="est")
sj_data
write_xlsx(sj_data, "master_files/model2/no2-2006.xlsx")
```

#### NO2 (2007)

```{r model-2-no2-2007, eval=eval_LC}
res_model_2_no2_2007 <- coxph(
  Surv(timeStudy, lung_cancer) ~ no2_2007per10 + ageBaseline + sex + householdIncomeCat + eduCat + bmiCat + smokingCat + race,
  data = df_pheno,
  ties = "efron"
)
summary(res_model_2_no2_2007)

sj_data = res_model_2_no2_2007 %>% sjPlot::get_model_data(type="est")
sj_data
write_xlsx(sj_data, "master_files/model2/no2-2007.xlsx")
```

#### NO2 (2010)
```{r model-2-no2-2010, eval=eval_LC}
res_model_2_no2_2010 <- coxph(
  Surv(timeStudy, lung_cancer) ~ no2_2010per10 + ageBaseline + sex + householdIncomeCat + eduCat + bmiCat + smokingCat + race,
  data = df_pheno,
  ties = "efron"
)
summary(res_model_2_no2_2010)

sj_data = res_model_2_no2_2010 %>% sjPlot::get_model_data(type="est")
sj_data
write_xlsx(sj_data, "master_files/model2/no2-2010.xlsx")

```

#### NO1 (2010)
```{r model-2-no1-2010, eval=eval_LC}
res_model_2_no1_2010 <- coxph(
  Surv(timeStudy, lung_cancer) ~ no1_2010per20 + ageBaseline + sex + householdIncomeCat + eduCat + bmiCat + smokingCat + race,
  data = df_pheno,
  ties = "efron"
)
summary(res_model_2_no1_2010)

sj_data = res_model_2_no1_2010 %>% sjPlot::get_model_data(type="est")
sj_data
write_xlsx(sj_data, "master_files/model2/no1-2010.xlsx")

```


## Citations

[@huang2021a; @tian2022a; @swanton2016]
